---
title: "Data types and structures, and Subsetting"
author: "Lucy Njoki"
date: "7/10/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objective of the discussion
- Data types and structures in R: Types and creating them

- How check type of the vector

- Missing values and other special values

- Subsetting data structures

# Data Types
- R has 6 data types namely: 

> character

> numeric

> logical

> integer

> complex

> raw

- Data structures are formed when elements of these data types combine e.g. atomic vectors(vector that contains data of single type).

- Examples include:

**character**: ```"b", "Kenya"```

**numeric:** ```3, 14.1```

**integer:** ```4L``` - ```L``` commands R to store the object as an integer

**complex:**```7+ 17i``` - contains the real and imaginary parts

**logical:**```True``` or ```False```

R has tools to check features of vectors or other objects

```class()``` - kind of object(high-level)

```typeof()``` - the data type of the object(low-level)

```length()``` - count of elements of the object

```attributes()``` - check if the object has any metadata; metadata refers to "data that provides information about other data". 

```{r}
y <- "Kenya"

typeof(y)
```

```{r}
attributes(y)
```

```{r}
p <- 1:20
p
typeof(p)
length(p)

q <- as.numeric(p) #converts to numeric
q
typeof(q)
```

# Data Structures

 - atomic vector
 
 - list
 
 - matrix
 
 - dataframe
 
 - factors
 
# Vectors

- A vector is the common and most basic structure of R. 

- Vectors are of two types:

> Atomic vectors

> Lists

## Different vector modes

Vector consists of element of mode of storage ```character```, ```logical```, ```integer``` or ```numeric```

Vectors can be created using ```vector()``` or ```c()```.

```{r}
vector() #default is logical

vector(mode = "character", length = 2)

character(length = 3)
logical(length = 3)
numeric(length = 3)
integer(length = 3)
```

or

```{r}
m <- c(1,2,3,4,5)
typeof(m)

n <- c(1L, 2L, 3L, 4L, 5L) 
typeof(n)
n <- as.integer(m)
typeof(n)

d <- c(TRUE, FALSE, TRUE)
typeof(d)
q <- c("Ke","Ug", "Tz")
typeof(q)
```
Let's examine the vectors
```{r}
typeof(n)
typeof(m)
length(q)
```

- Vectors can also be created as a sequence of numbers.

```{r}
l <- 1:20
l
seq(20)

seq(from = 1, to = 10, by = 0.5)
seq(1,10, by = 0.5)
```

## Missing Data

- Missing data in vectors are represented as ```NA```

```{r}
x <- c(1,2, 4, 0.9, NA)
x
is.na(x)
sum(is.na(x))
complete.cases(x)
anyNA(x)

```

## Other special Values

```inf```  - infinity (+ve or-ve)

```{r}
1/0
#1/Inf
```

```NaN``` - Not a Number(Undefined value)
 
```{r}
sp <- 0/0
sp
is.nan(sp) #tests for NaN
sum(is.nan(sp))
```

## Vectors wth mixed data types

```{r}
a <- c(1.7, "a")
b <- c(TRUE, 2)
e <- c("a", TRUE)

class(a)
class(b)
class(e)
```

Note:
- When R converts the mode of storage based on its content, it is referred to as __implicit coercion__

1. A vector consisting of numeric and character elements, LCM which is picked as the class of the vector is character.

2.  A vector consisting of numeric and logical, LCM which is picked as the class of the vector is numeric.

3. A vector consisting of logical and character, LCM which is picked as the class of the vector is character.

- However, we can convert elements between different modes of storage is called __explicit coercion__

```as.<class_name>()```

```{r}
typeof(as.numeric(2))
```

## Subsetting

- Elements of a vector are obtained using []

```{r}
x <- c(1,2,3,4,7,10)
x[5]
x[c(5,1,3)]
```

# Lists

- Elements are of different data types.

```list()```

Vectors can also be coerced into lists ```as.list()```

```{r}
t <- list(1, "a", TRUE, 1+4i)
class(t)
```

## Subsetting Lists

- Elements are obtained  using [[]]

```{r}
mylist <- list(a = "name", b = 1:10, data = head(iris))
mylist
names(mylist)

vec <- 1:10
vec <- as.list(vec)

class(vec)
```
```{r}
t[[1]]
t[[3]]
#t[1] #Wrong
mylist$b
mylist$a
```

# Matrices

- Matrices are vectors with a dimension attribute

Can be created using ```matrix()```, ```cbind()``` or ```rbind()```

```{r}
ma <- matrix(nrow = 2, ncol = 3, data = c(1,2,3,4,5,6))
ma

dim(ma) # no. of rows and columns

attributes(ma) # dim; metadata

class(ma)

typeof(ma)

```

- Constructed column-wise that is entries are starting upper left, running down.

```class()``` shows that ma is a matrix while ```typeof()``` shows that basically the matrix is a numeric integer.

## Other ways of creating matrices

```{r}
va <- 1:10
dim(va) <- c(2,5)
va
```

- Bind rows or columns using ```rbind()``` or ```cbind()``` respectively.

```{r}
s <- 1:3
t <- 10:12
p <- 50:52

cbind(s,t,p)
```
```{r}
rbind(s,t,p)
```

```byrow``` argument specifies how the matrix is filled.

```{r}
vb <- matrix(data = 1:20, nrow = 2, ncol = 5, byrow = TRUE)
vb

vb[2,3]
vb[1,5]
vb[,1] #specific column including all the rows
vb[1,] #specific row including all the columns
```

# Factors

- Can be ordered or unordered.

```factor()```

```{r}
xfac <- factor(c("yes", "no"))
xfac
class(xfac)

#to remove levels leaving it as a character
unclass(xfac)
```



```{r echo=FALSE, results=  'asis'}
#working progress
library(knitr)
x <- c("Dimensions","1-D", " 2-D")
y <- c("Homogenous","atomic vector" ,"matrix")
z <- c("Heterogeneous","list","data frame")
kable(list(x,y,z), format = "html", row.names = FALSE, caption = "Summary of one-dimensional and two-dimensional data structures in R in relation to diversity of data types they can contain.")
```
- Homogeneous data structures: can only store a single type of data (numeric, integer, character, etc.).

- Heterogeneous data structures: can store more than one type of data at the same time.

*_Source: https://swcarpentry.github.io/r-novice-inflammation/13-supp-data-structures/ & _*
*_Data Science foundation using R: intro to R(Coursera)_*